Preregistration: Changes

### 2023-01-30 manuscript.Rmd
- major updates to preprint

abstract          : "In times of crisis such as the Corona pandemic citizens need to stay informed about recent events, the latest political decisions, or mandatory protection measures. To this end, many people use various types of media, and increasingly social media. However, because social media are particularly engaging, some find it hard to disconnect and cannot stop 'doomscrolling'. In this preregistered study, I investigate whether using social media for COVID-19 related reasons affects psychological well-being. To answer this question I analyzed data from the Austrian Corona Panel Project, which consists of 24 waves with overall 3,018 participants. I ran three random effects within between models, controlling for several stable and varying confounders. Results showed that the effects of COVID-19 related social media use on well-being were very small, arguably too small to matter. The findings suggest that fears that social media use during times of crisis impairs well-being are likely to be unfounded."
keywords          : "COVID-19, Coronavirus, well-being, affect, life satisfaction, social media use, news use, communication, random effects within between model, panel study, longitudinal."

@@

floatsintext      : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : yes
draft             : no

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
# output            : papaja::apa6_word
header-includes:
  - \setlength{\parskip}{0em}
  - \raggedbottom
  - \note{\clearpage}
---

```{r analysis-preferences, include=F, cache=F}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed,
                      cache=T, echo=F
                      )
```

@@

# load packages
packages <- c("devtools", "english", "ggplot2", "kableExtra", 
              "lavaan", "lme4", "lmerTest",
              "magrittr", "semPlot", "tidyverse")
lapply(c(packages, "papaja", "td"), library, character.only = TRUE)

@@

During the COVID-19 pandemic, 
numerous events unfolded in quick succession and several open questions emerged.
How dangerous is the virus? 
Is it spreading in my region? 
How is it transmitted, and how can I protect myself?
Because for many it was (and still is) a matter of life or death, people aimed to stay informed regarding the latest developments.
Governments around the world implemented safety measures, such as wearing masks, keeping physical distance, or enforcing lockdowns.
In this extraordinary situation, many people used media excessively to attain information, and especially social media were at an all time high [@statistaAverageDailyTime2021].

Some people actually couldn't stop using social media to learn about COVID-19 related news.
A new phenomenon emerged, termed "doomscrolling":
Users were glued to their screens and found it hard to pursue other relevant activities such as working, taking a break, or looking after their children [@kleinDarklySoothingCompulsion2021].
As doomscrolling increased it became doubtful whether such social media is helpful, or whether it created an additional burden on mental health [@sandstromDoomscrollingCOVIDNews2021].
These concerns seem justified: 
A study with 6,233 people from Germany conducted during the pandemic found that "[f]requency, duration and diversity of media exposure were positively associated with more symptoms of depression" [@bendauAssociationsCOVID19Related2021, p. 283].

As a result, with this study I want to build on this research and investigate whether or not COVID-19 related social media use affected well-being during the pandemic.
To this end, I analyzed a large-scale panel study from the Austrian Corona Panel Project [@kittelAustrianCoronaPanel2020]. 
The panel consists of 24 waves and has an overall sample size of `r nrow(d_wide) %>% prettyNum(big.mark = ",")` participants.
<!-- At least 1,500 participants took part per wave, and the sample is representative of the Austrian population. -->
The panel study collected a large number of psychological and demographic variables.
<!-- Being able to control for many confounding third variables, both stable and varying, together with the longitudinal design, this creates a unique opportunity  -->
I explicitly aimed to investigate the causal effects of COVID-19 related social media use on well-being.

## Defining Well-being and Media Use

The underlying theories that guided the selection of variables for my analysis are the two-continua model of mental health [@greenspoonIntegrationSubjectiveWellbeing2001] and the hierarchical taxonomy of computer-mediated communication [@meierComputermediatedCommunicationSocial2020a].
According to the two-continua model, mental health consists of (a) psychopathology and (b) well-being.
<!-- Because the aim of this study is better to understand typical users and everyday contexts, my focus will be on well-being. -->
Well-being can be differentiated into subjective and psychological well-being [@dienerAdvancesOpenQuestions2018].
Whereas subjective well-being emphasizes hedonic aspects such as happiness and joy, psychological well-being addresses eudaimonic aspects such as fulfillment and meaning.
Subjective well-being is primarily about achieving positive affect and avoiding negative affect.
One of the most prominent indicators of well-being is life satisfaction.
In my view, because it represents a general appraisal of one's life, life satisfaction is best thought of as a meta concept that combines psychological and subjective well-being.
Notably, life satisfaction is stable and fluctuates only little, whereas it's the exact opposite for affect [@dienlinImpactDigitalTechnology2020].
<!-- Everyone can feel bad on any given day, but whether or not one feels generally comfortable in one's life rather changes across years. -->
To capture well-being in this study I thus build on life satisfaction, positive affect, and negative affect.
Together, this should provide an encompassing perspective on potential media effects.

The hierarchical taxonomy of computer-mediated communication differentiates six levels of how people engage with digital technology.
First, the device (e.g., smartphone); second, the type of application (e.g., social networking site); third, the branded application (e.g., Twitter); fourth, the feature (e.g., status post); fifth, the interaction (e.g., one-to-many); and sixth, the message (e.g., content) [@meierComputermediatedCommunicationSocial2020a].
Whereas the first four levels focus on the communication _channel_, the last two address the communication _type_.
To measure social media use for the consumption of COVID-19 related news and topics, I here employ both the channel and the type of communication perspective, which together provides a nuanced understanding of communication.

First, I investigate how different types of communication affect well-being.
Specifically, I differentiate between active and passive use.
I distinguish (a) _reading_ (passive), (b) _posting_ (active), and (c) _liking and sharing_ COVID-19 related posts (both active and passive).
Second, I analyze how using the most prominent branded applications affects well-being, and whether this effect changes across applications.
Branded apps are separate entities with potentially divergent effects. 
Twitter might have a different effect as compared to WhatsApp because of their respective affordances.
For example, @waterlooNormsOnlineExpressions2018 found that it's more adequate to express negative emotions on WhatsApp than on Twitter or on Instagram. 
The branded applications investigated here are Facebook, Twitter, Instagram, WhatsApp, and YouTube.
Worth noting, this study is not about _general_ social media use during times of COVID, but on social media use _focused_ on COVID-19 related content. 
Examples of such media use include posting thoughts about the pandemic or retweeting COVID-19 related news.

## Effects of Social Media on Well-Being

So far, there is only little empirical research on COVID-19 related social media use on well-being.
In their study on the relations between media use and mental health during the pandemic, @bendauAssociationsCOVID19Related2021 found that people who used social media as a primary source of information reported on average "significantly more unspecific anxiety and depression [] and significantly more specific COVID-19 related anxiety symptoms" (p. 288).
@edenMediaCopingCOVID192020 analyzed the media use of 425 US college students during the first wave of the pandemic, finding both positive and negative relations with well-being.
In a sample of 312 respondents collected via Amazon Mechanical Turk, @choiMediatedCommunicationMatters2021 reported that people who used media to attain information were more lonely and less satisfied with their lives.
@stainbackCOVID1924News2020 analyzed a large-scale study with 11,537 respondents from the US and found that increased COVID-19 media consumption was related to more psychological distress.
Together, the literature emphasizes potentially negative effects of social media as news use [see also @liuRelationOfficialWhatsAppdistributed2020; @riehmAssociationsMediaExposure2020].
However, note that all of these findings represent between-person relations stemming from cross-sectional data (see below).
We therefore don't know whether the differences in mental health and well-being are due to social media use or due to other third variables, such as age, health, employment, or education.

The question of whether and how social media use affects well-being _in general_, on the other hand, is well-researched.
This also holds true for the different types of communication such as active or passive use.
A meta review (i.e., an analysis of meta-analyses) found that the relation between social media use and well-being is likely in the negative spectrum but very small [@meierComputermediatedCommunicationSocial2020a]---potentially too small to matter.
What determines whether or not an effect is considered small or trivial?
As a starting point, we could refer to standardized effect sizes.
According to @cohenPowerPrimer1992, small effect sizes start at _r_ = .10.
And indeed, several if not most of the current meta-analyses find effect sizes below that threshold [@huangTimeSpentSocial2017; @meierComputermediatedCommunicationSocial2020a; @fergusonThisMetaanalysisScreen2021].

These overviews are well aligned with several individual studies employing advanced methods [@orbenSocialMediaEnduring2019; @przybylskiDoesTakingShort2021a; @schemerImpactInternetSocial2021; @kerestesAdolescentsOnlineSocial2020].
For example, @beyensSocialMediaUse2021 reported that although for some users (roughly one quarter) the effects of social media use on well-being were negative, for almost the same number of users they were positive, while for the rest the effects were neutral.
In conclusion, most effects are likely somewhere between trivial and small.
I therefore expect that also in the case of COVID-19 related social media use effects will be trivial to small.

From a theoretical perspective, how could we explain whether COVID-19 related social media use might affect well-being?
In what follows, I outline potential arguments as to why the effect might be positive or negative, direct or indirect, or nonexistent.
In advance, there doesn't seem to be a clear winner, and it's likely that both positive and negative effects cancel each other out.

First, one could assume a _direct_ negative effect on well-being, and especially on positive or negative affect, which are more volatile and fluctuating.
Dangers, inequalities, corruption---these were the headlines during the pandemic across many countries worldwide.
If one learns about such events, the initial reaction might be shock, fear, or dismay.
Repeatedly consuming such news can be depressing, perhaps even changing some general perspectives on life.
<!-- So, just like being hit by a hammer hurts and we don't need any "mediating mechanism", this could be the case here as well. -->
That said, because not all news was negative, and because many people showed solidarity and compassion, there was also positive and uplifting content, potentially compensating for the negative effects.
<!-- However, in light of a worldwide pandemic with millions of deaths, the negative direct effect seems more plausible. -->

There could also be _indirect_ effects.
When doomscrolling, users are captivated to such an extent that they cannot stop using social media.
For example, during the pandemic social media use was at an all-time high in the US [@statistaAverageDailyTime2021].
In general, as has been expressed by many before, it is most likely that moderate social media use is not detrimental [@orbenTeenagersScreensSocial2020].
Overuse, however, might be more critical, and several studies have shown more pronounced negative effects for extreme users [@przybylskiLargescaleTestGoldilocks2017].
To explain, overuse likely impairs well-being if it replaces meaningful or functional activities such as meeting others, working, actively relaxing, or exercising.
So if a society collectively overuses social media when doomscrolling, there is potential for negative effects.
<!-- If overuse replaces such activities it's reasonable to assume that it's also detrimental. -->

On the other hand, one can make the case that overuse might also be beneficial, especially in times of a pandemic---even if the use is mainly COVID-19 related.
Exchanging COVID-19 related messages with friends via WhatsApp might replace the in-person contact one would have otherwise, but which is literally impossible at the time.
In situations where meaningful and functional activities are prohibited, using social media to exchange about COVID-19 related topics might not be the worst idea.
Besides, given that nowadays a large number of experts, scientists, and politicians converse directly on social media, one can get first-hand high quality information on current developments.
<!-- On the other hand, there is of course also much disinformation, and "bingeing" on COVID-19 fake news might also pose risks for impaired well-being. -->

Together, the strongest argument to me is that _in general_ the effects of social media on well-being are, on average, small at best.
Because this study only looks at _one part_ of social media use---namely, COVID-19 related interactions---it is very focused, diminishing the overall potential of the effects even further.
Whether or not using social media for COVID-19 related aspects is detrimental during a pandemic is also not entirely clear.
<!-- At least, there does not seem to be a case where we should expect a clear deviation to the negative, despite anecdotes of doomscrolling. -->
Therefore, I expect that COVID-19 related communication on social media doesn't affect well-being in a meaningful or relevant way.

@@

# Current Study
## Smallest Effect Size of Interest

Testing this hypothesis, however, is not trivial.
First, in contrast to most hypotheses typically posited in the social sciences it implicitly contains an effect size, a so-called smallest effect size of interest (SESOI).
Effectively testing this hypothesis necessitates defining what's considered a "trivial effect size" and what's not.
Above I already referred to standardized effect sizes.
However, standardized effect sizes should only be a first step toward evaluating an effect's relevance [@baguleyStandardizedSimpleEffect2009].
Standardized effect sizes are determined by a sample's variance,[^cohensd] which is problematic: 
The question of whether or not social media use affects a particular person in a relevant way should not depend on the variance in the sample in which that person's data were collected.
Instead, it should depend on absolute criteria.

What could be a minimally interesting, nontrivial effect?
Because this is a normative and ultimately philosophical question, there can never be a clear, single, or unanimous answer. 
<!-- In the end, it is a personal question. -->
However, it is still necessary and helpful to try provide such a plausible benchmark. 
I therefore suggest the following SESOI for this research question:

[^cohensd]: Consider the effect size Cohen's _d_: The mean's of the two groups that are to be compared are subtracted from one another and then divided by the sample's standard deviation [@cohenPowerPrimer1992]. Hence, if there is more deviation/variance in a sample, the effect size decreases, even if the difference of the group's means stays the same.

>>>>

note: |
  Preprint submitted for publication. Please cite carefully.

abstract          : "In times of crisis such as the Corona pandemic citizens need to stay informed about recent events, the latest political decisions, or mandatory protection measures. To this end, many people use various types of media, and increasingly social media. However, because social media are particularly engaging, some find it hard to disconnect and cannot stop 'doomscrolling'. In this preregistered study, I investigate whether using social media for COVID-19 related reasons affects psychological well-being. To answer this question I analyzed data from the Austrian Corona Panel Project, which consists of 24 waves with overall 3,018 participants. I ran three random effects within between models, controlling for several stable and varying confounders. Results showed that the effects of COVID-19 related social media use on well-being were very small, arguably too small to matter. The findings suggest that fears that social media use during times of crisis impairs well-being are likely to be unfounded."
keywords          : "COVID-19, well-being, social media, news use, panel study."

@@

floatsintext      : yes
endnotes          : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no

header-includes:
  - |
    \makeatletter
    \renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
      {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
      {-1em}%
      {\normalfont\normalsize\bfseries\typesectitle}}

    \renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
      {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
      {-\z@\relax}%
      {\normalfont\normalsize\bfseries\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
    \makeatother
    \usepackage{endnotes}
    \setlength{\parskip}{0em}
    \raggedbottom
    \note{\clearpage}

csl               : "`r system.file('rmd', 'apa7.csl', package = 'papaja')`"
documentclass     : "apa7"
output            :
  papaja::apa6_pdf:
    latex_engine  : xelatex
# output            : papaja::apa6_docx
# output            : html_document
---

```{r analysis-preferences, include=F, cache=F}
set.seed(170819) # Seed for random number generation
knitr::opts_chunk$set(cache.extra = knitr::rand_seed,
                      cache=T, echo=F)
```

@@

# load packages
packages <- c("devtools", "english", "GGally", "ggplot2", "gridExtra", "kableExtra", 
              "lavaan", "lme4", "lmerTest",
              "magrittr", "papaja", "tidyverse")
lapply(c(packages, "td"), library, character.only = TRUE)

@@

During the COVID-19 pandemic it was critical to stay informed regarding the latest developments.
How dangerous is the virus? 
In what region is it spreading? 
How is it transmitted? What are the current safety regulations?
To obtain relevant information, many people heavily relied on social media, which were at an all time high [@statistaAverageDailyTime2021].

Some actually could not stop using social media to learn about COVID-19 related news.
A new phenomenon termed "doomscrolling" emerged.
Many users were glued to their screens and found it hard to pursue other relevant activities such as working, taking a break, or even looking after their children [@kleinDarklySoothingCompulsion2021].
In the media it was hence increasingly asked whether using social media for COVID-19 related reasons would, next to all other stressors, create an additional burden on mental health [@sandstromDoomscrollingCOVIDNews2021].
Although research has begun addressing this question [@bendauAssociationsCOVID19Related2021; @bradleyStressMoodSmartphone2021; @choiMediatedCommunicationMatters2021; @dornemannHowGoodBad2021; @edenMediaCopingCOVID192020; @guazziniSecondWaveAnalysis2022; @latikkaLonelinessPsychologicalDistress2022; @liuRelationOfficialWhatsAppdistributed2020; @riehmAssociationsMediaExposure2020; @sewallObjectivelyMeasuredDigital2021; @stainbackCOVID1924News2020; @yuePassiveSocialMedia2022], it still largely unknown if COVID-19 related social media use during the pandemic has had a meaningful impact on well-being.
By analyzing large-scale panel data from Austria and by focusing on intra-individual causal effects, this study aims to answer to this question.

## Understanding Well-being and Media Use

<!-- Two underlying theories guided the selection of variables for this study, namely the two-continua model of mental health [@greenspoonIntegrationSubjectiveWellbeing2001] and the hierarchical taxonomy of computer-mediated communication [@meierComputermediatedCommunicationSocial2020a]. -->
<!-- According to the two-continua model, mental health consists of (a) psychopathology and (b) well-being. -->
<!-- <!-- Because the aim of this study is better to understand typical users and everyday contexts, my focus will be on well-being. -->
<!-- Well-being can be differentiated into subjective and psychological well-being [@dienerAdvancesOpenQuestions2018]. -->
<!-- Whereas subjective well-being emphasizes hedonic aspects such as happiness and joy, psychological well-being addresses eudaimonic aspects such as fulfillment and meaning. -->
<!-- Subjective well-being is primarily about achieving positive and avoiding negative affect. -->
<!-- One of the most prominent indicators of well-being is life satisfaction. -->
<!-- In my view, because it represents a general appraisal of one's life, life satisfaction is best thought of as a meta concept combining psychological and subjective well-being. -->
<!-- Notably, life satisfaction is stable and fluctuates only little, whereas it's the exact opposite for affect [@dienlinImpactDigitalTechnology2020]. -->
<!-- <!-- Everyone can feel bad on any given day, but whether or not one feels generally comfortable in one's life rather changes across years. -->
<!-- To capture well-being in this study I thus build on life satisfaction, positive affect, and negative affect. -->
<!-- Together, this should provide an encompassing perspective on potential media effects. -->

<!-- The hierarchical taxonomy of computer-mediated communication differentiates six levels of how people engage with digital technology. -->
<!-- First, the device (e.g., smartphone); second, the type of application (e.g., social networking site); third, the branded application (e.g., Twitter); fourth, the feature (e.g., status post); fifth, the interaction (e.g., one-to-many); and sixth, the message (e.g., content) [@meierComputermediatedCommunicationSocial2020a]. -->
<!-- Whereas the first four levels focus on the communication _channel_, the last two address the communication _type_. -->
<!-- Distinguishing different communication channels and communication types is important, because the effects on well-being likely differ across communication channels and communication types. -->
<!-- Whereas active social media use such as chatting is routinely linked to improved well-being, passive use such as reading is often considered more negative [@dienlinImpactDigitalTechnology2020]. -->
<!-- Similarly, branded apps are separate communication entities with potentially divergent effects and affordances. -->
<!-- For example, @waterlooNormsOnlineExpressions2018 found that it's more adequate to express negative emotions on WhatsApp than on Twitter or on Instagram. -->
<!-- Especially during a pandemic, it makes sense to analyze if users engage with COVID-19 related content on Instagram, where communication is more positive, or on Facebook, where communication is more critical. -->
<!-- First studies suggest that during the pandemic Instagram use was indeed more beneficial for well-being than Facebook use [@masciantonioDonPutAll2021]. -->
<!-- In this study, to measure the effects of social media use focused on COVID-19 related news and topics, I adopt both the channel and the type of communication perspective. -->
<!-- Together, this should offer a nuanced and comprehensive understanding of communication. -->

This study investigates how different _facets_ of well-being are affected by different _types_ and different _channels_ of communication [@meierComputermediatedCommunicationSocial2020a].
<!-- , namely active and passive use [@ellisonWhyWeDon2020; @meierComputermediatedCommunicationSocial2020a]. -->
<!-- Defining what constitutes active and what passive use is not always clear, and different understandings are currently discussed [@ellisonWhyWeDon2020; @meierComputermediatedCommunicationSocial2020a]. -->
<!-- Reading is generally considered as passive and writing as active use, while there are also specific behaviors falling somewhere in-between such as liking or sharing content [@meierDoesPassiveSocial2022]. -->
<!-- In this study,  -->
Building on the typology of subjective well-being [@dienerAdvancesOpenQuestions2018], three different well-being facets are analyzed: life satisfaction, positive affect, and negative affect.
Because effects of social media depend on how they are used [@verduynPassiveFacebookUsage2015], I further distinguish three types of use and five popular channels. 
The types of use include reading, posting, and liking and sharing COVID-19 related posts.
In doing so, this study analyzes social media use focused on COVID-19 related content, which includes posting thoughts about the pandemic, reading posts and comments, or retweeting COVID-19 related news.
The five channels to be investigated are Facebook, Twitter, Instagram, WhatsApp, and YouTube, which at the time ranked among the most popular social media services in Austria.

<!-- ## Theorizing Social Media Effects on Well-Being -->

<!-- <!-- In conclusion, most effects are likely somewhere between trivial and small. -->
<!-- <!-- I therefore expect that also in the case of COVID-19 related social media use effects will be trivial to small. -->

How could the various types and channels of COVID-19 related social media use affect well-being?
<!-- In what follows, I outline potential arguments as to why the effect might be positive or negative, direct or indirect, or nonexistent. -->
According to the set-point theory, well-being is surprisingly stable [@lykkenHappinessWhatStudies1999].
Although specific events such as marriage or salary can have significant impacts on well-being, in most cases effects are only short-term, with well-being after some time routinely returning to prior levels [@sheldonStabilityHappinessTheories2014].
Only very specific factors such as unemployment, disability, or death can cause long-term changes in well-being [@lucasAdaptationSetpointModel2007].
<!-- So although well-being can change this does not happen easily. -->
<!-- Although the set-point theory has been criticized lately because events such as marriage, unemployment, or death can indeed lead to long-term changes, it is still widely maintained that well-being is very stable and hence hard to be influenced  -->
Can media use be such a factor?
Current literature overviews suggest no: 
Although social media use on average decreases well-being, the effects are small [@meierComputermediatedCommunicationSocial2020a].
In addition, the effects of media use differ across individuals and types of content [@valkenburgDifferentialSusceptibilityMedia2013].
Whereas for some users social media are more beneficial, for others they are more harmful.
For example, in one study it was estimated that roughly one quarter of all users experienced negative effects, another quarter positive effects, while for the rest the effects were neutral [@beyensSocialMediaUse2021].
This finding is aligned with the Differential Susceptibility to Media Effects Model: 
<!-- This finding is aligned with the Differential Susceptibility to Media Effects Model:  -->
Although there is substantial _variation_ of media effects for individual users, there are both negative and positive aspects, leading to small effects on average [@valkenburgDifferentialSusceptibilityMedia2013]---as confirmed by several literature overviews [@dienlinImpactDigitalTechnology2020; @huangTimeSpentSocial2017; @meierComputermediatedCommunicationSocial2020a; @orbenTeenagersScreensSocial2020].

Why are the effects of social media use on well-being small on average?
Using Social media is ambivalent.
It can impair well-being when causing embarrassment, stress, or disinformation, and it can improve well-being when providing connectedness, information, or entertainment [@buchiDigitalWellbeingTheory2021].
Although social media are often associated with negative outcomes, there are a large number of positive effects, which either offset or at least significantly reduce negative outcomes.
Positive outcomes encompass finding relevant information; maintaining and fostering relationships; expressing one's personality; or entertaining oneself [e.g., @pelletierOneSizeDoesn2020].
Negative effects include distraction; displacement of other meaningful activities; consumption of shallow, misleading or challenging content; or negative social comparisons [e.g., @meierDoesPassiveSocial2022].
People are much more likely to use media that offer plentiful benefits [@katzUsesGratificationsResearch1973].
Users implicitly learn what media help them regulate their mood and thereby well-being [@zillmannMoodManagementCommunication1988].
Given the ubiquitous use of social media, the many benefits they offer, plus the general and implicit competence to use media that foster mood management, it is hence unlikely that the effects of social media use on well-being are profusely negative.

<!-- Two prominent media effect theories argue (mostly implicitly) against strong average negative effects. -->
<!-- According to mood management theory [@zillmannMoodManagementCommunication1988], using media can substantially affect people's moods. -->
<!-- Use can be stimulating or overwhelming, relaxing or boring. -->
<!-- After some time, users implicitly learn what media help them balance their mood and affect according to their own situational needs [@zillmannMoodManagementCommunication1988]. -->
<!-- Those media that eventually become part of one's media repertoire are hence, on average, beneficial for users to regulate their mood. -->
<!-- Using experience sampling of well-being and logs of social media use, a study with 82 participants from Italy found that after episodes of social media use, levels of positive affect increased significantly [@marcianoDynamicsAdolescentsSmartphone2022]. -->

<!-- While mood management theory considers media use mainly driven by implicit learning experiences, uses and gratifications theory upholds that the process is more explicit and rational [@katzUsesGratificationsResearch1973]. -->
<!-- Users select those media that they expect to have a desired effect, for example on mood, knowledge, or entertainment.  -->
<!-- If those beneficial media effects do not exist or if they are not expected, people will spend their time elsewhere. -->
<!-- And social media offer several beneficial effects, explaining why they are used that much. -->
<!-- They help find relevant information, maintain and foster relationships, express one's personality, and entertain oneself [@pelletierOneSizeDoesn2020]. -->
<!-- In conclusion, because people spend so much time on social media consuming COVID-19 related content, according to both mood management theory and uses and gratifications theory this is indirect proof that average effects on well-being are likely not particularly negative. -->

<!-- But people can also misjudge media effects and are often overly optimistic [@metzgerComparativeOptimismPrivacy2017]. -->
<!-- Precisely because social media have so many positive consequences, one can ask if this is not where the actual problem lies. -->
<!-- In other words, social media might not problematic because they are inherently bad, but rather because they are too good. -->
<!-- As with many other things, there can be too much of a good thing. -->
<!-- It is therefore often asked whether social media can become addictive, and users sometimes express this fear themselves [@yangCanWatchingOnline2021].  -->
<!-- However, a recently published meta-analysis found that the two most prominent measures of addiction, the Bergen Facebook Addiction Scale and the Bergen Social Media Addiction Scale, have only small relations to well-being [@duradoniWellbeingSocialMedia2020].  -->
<!-- In addition, the general idea of labeling excessive social and new media use as addiction was criticized, arguing that social media represent new regular behaviors that should not be pathologized [@galerHowMuchToo2018; @vanrooijWeakScientificBasis2018]. -->

<!-- Because media effects can differ across users, situations, and content [@livingstoneEuropeanResearchChildren2018; @valkenburgDifferentialSusceptibilityMedia2013], I now briefly focus on the effects of COVID-19 related social media use specifically. -->
<!-- First, one could assume a _direct_ negative effect on well-being, and especially on positive or negative affect, which are more volatile and fluctuating. -->
<!-- Dangers, inequalities, corruption---these were the headlines during the pandemic across many countries worldwide. -->
<!-- If one learns about such events, the initial reaction might be shock, fear, or dismay. -->
<!-- Consuming such news can be depressing [@dornemannHowGoodBad2021], perhaps even changing some general perspectives on life. -->
<!-- <!-- So, just like being hit by a hammer hurts and we don't need any "mediating mechanism", this could be the case here as well. -->
<!-- That said, because not all news was negative, and because many people showed solidarity and compassion, there was also positive and uplifting content, potentially compensating for the negative effects [@dornemannHowGoodBad2021]. -->
<!-- A study with 2.057 respondents from Italy reported that during the pandemic virtual community and social connectedness even increased [@guazziniSecondWaveAnalysis2022]. -->

<!-- In a study with 735 participants from Finland, levels of loneliness did not decrease during the pandemic, and people who engaged more on social media experienced less loneliness [@latikkaLonelinessPsychologicalDistress2022]. -->
<!-- However, in light of a worldwide pandemic with millions of deaths, the negative direct effect seems more plausible. -->

<!-- Second, there could also be _indirect_ effects. -->
<!-- Although it is most likely that moderate social media use is not detrimental [@orbenTeenagersScreensSocial2020], overuse, however, might be more critical, and several studies have shown more pronounced negative effects for extreme users [@przybylskiLargescaleTestGoldilocks2017]. -->
<!-- To explain, overuse could impair well-being if it replaces meaningful or functional activities such as meeting others, working, actively relaxing, or exercising. -->
<!-- Another potentially negative mechanism at play are problematic social comparison processes.  -->
<!-- Therefore, if a society collectively overuses social media during a pandemic, there is potential for negative effects. -->
<!-- If overuse replaces such activities it's reasonable to assume that it's also detrimental. -->

<!-- On the other hand, one can make the case that using social media for COVID-19 related reasons might even be beneficial, especially in times of a pandemic. -->
<!-- Exchanging COVID-19 related messages with friends via WhatsApp might replace the in-person contact one would have otherwise, but which is literally impossible at the time. -->
<!-- In situations where meaningful and functional activities are prohibited, using social media to exchange about COVID-19 related topics might not be the worst idea. -->
<!-- Besides, given that nowadays a large number of experts, scientists, and politicians converse directly on social media, one can get first-hand high quality information on current developments. -->
<!-- <!-- On the other hand, there is of course also much disinformation, and "bingeing" on COVID-19 fake news might also pose risks for impaired well-being. -->

<!-- To summarize, it seems that from a theoretical perspective it is most likely that the average effects of social media use on well-being are negligible.  -->
<!-- Building on established theories from Communication and current empirical findings, we would not assume that effects are either profoundly negative or strongly positive. -->

<!-- ## Empirical Studies on Social Media Effects -->

There is still little empirical research on how well-being is affected by social media use that is focused on COVID-19 specifically.
Results are mixed.
When browsing social media for COVID-19 related news, many users reported being captivated to such an extent they could not stop using social media [@kleinDarklySoothingCompulsion2021].
<!-- During the pandemic social media use was at an all-time high in the US [@statistaAverageDailyTime2021], for example to cope with the required physical distancing measures. -->
In a study with 1,131 residents from Wuhan in China, people who spent more time in quarantine also spent more time on social media [@yuePassiveSocialMedia2022]. 
Those who spent more time on social media also engaged in more upward social comparison, which was related to increased levels of stress.
People who used social media as a primary source of information reported on average "significantly more unspecific anxiety and depression [...] and significantly more specific COVID-19 related anxiety symptoms" [@bendauAssociationsCOVID19Related2021, p. 288].
@edenMediaCopingCOVID192020 analyzed how 425 US college students used media during the first wave of the pandemic, finding both positive and negative relations with well-being.
<!-- In a sample of 312 respondents collected via Amazon Mechanical Turk, people who used media to attain information were more lonely and less satisfied with their lives [@choiMediatedCommunicationMatters2021]. -->
In a large-scale study with 11,537 respondents from the US, increased COVID-19-related media consumption was related to more psychological distress [@stainbackCOVID1924News2020].
A four-wave panel study with 384 young adults from the U.S. analyzed the effects of general digital technology use during the pandemic on mental health, finding that digital technology did not have significant effects on mental health [@sewallObjectivelyMeasuredDigital2021; for a similar study with comparable results, see @bradleyStressMoodSmartphone2021].
A study with 2.057 respondents from Italy reported that during the pandemic virtual community and social connectedness increased [@guazziniSecondWaveAnalysis2022].
In a study with 735 participants from Finland, levels of loneliness were stable during the pandemic, and people who engaged more on social media experienced less loneliness [@latikkaLonelinessPsychologicalDistress2022].
Together, the literature is mixed, showing both positive and negative affects of social media use focused on or during COVID-19 on well-being [see also @dornemannHowGoodBad2021; @liuRelationOfficialWhatsAppdistributed2020; @riehmAssociationsMediaExposure2020].
<!-- However, note that all of these findings represent between-person relations stemming from cross-sectional data (see below). -->
<!-- We therefore don't know whether the differences in mental health and well-being are due to social media use or due to other third variables, such as age, health, employment, or education. -->

<!-- The question of whether and how social media use affects well-being _in general_, on the other hand, is well-researched. -->
<!-- This also holds true for the different types of communication such as active or passive use. -->
<!-- What determines whether or not an effect should be considered small or trivial? -->
<!-- As a starting point, we could refer to standardized effect sizes. -->
<!-- According to @cohenPowerPrimer1992, small effect sizes start at _r_ = .10. -->
<!-- Finally, several individual studies employing advanced methods found smalls relations between social media use and well-being [@orbenSocialMediaEnduring2019; @przybylskiDoesTakingShort2021a; @schemerImpactInternetSocial2021; @kerestesAdolescentsOnlineSocial2020]. -->
<!-- Although for roughly one quarter of all users the effects of social media use on well-being were negative, for almost the same number of users they were positive, while for the rest the effects were neutral [@beyensSocialMediaUse2021]. -->
<!-- This finding is aligned with the Differential Susceptibility to Media Effects Model:  -->
<!-- Although there is substantial _variation_ of media effects for individual users, the _average_ effects reported in the literature are often small [@valkenburgDifferentialSusceptibilityMedia2013].  -->

<!-- Together, the strongest argument to me is that _in general_ the effects of social media on well-being are, on average, small at best.  -->
<!-- Because this study only looks at _one part_ of social media use---namely, COVID-19 related interactions---it is very focused, diminishing the overall potential of the effects even further. -->
<!-- Whether or not using social media for COVID-19 related aspects is detrimental during a pandemic is also not entirely clear. -->
<!-- At least, there does not seem to be a case where we should expect a clear deviation to the negative, despite anecdotes of doomscrolling. -->

In conclusion, these mixed empirical results, together with the observation that social media effects on well-being are very small in general, I expect that COVID-19 related communication on social media does not affect well-being in a meaningful or relevant way.

<!-- A meta review (i.e., an analysis of meta-analyses) found that the relation between social media use and well-being is likely in the negative spectrum but very small, potentially too small to matter [@meierComputermediatedCommunicationSocial2020a]. -->
<!-- And indeed, several if not most of the current meta-analyses find small to trivial effects [@huangTimeSpentSocial2017; @meierComputermediatedCommunicationSocial2020a; @fergusonThisMetaanalysisScreen2021]. -->

@@

<!-- In this study, this general hypothesis will be analyzed specifically for the three communication types of (a) time spent reading, (b) liking and sharing, and (c) actively posting COVID-19 related content. -->
<!-- In addition, I will analyze how well-being is influenced by spending time on five prominent social media apps, including (a) Facebook, (b) Instagram, (c) Twitter, (d) WhatsApp, and (e) YouTube.  -->

# Method
## Preregistration

The hypotheses, the sample, the measures, the analyses, and the inference criteria (SESOI, _p_-value) were preregistered on the Open Science Framework, accessible here: https://osf.io/87b24/?view_only=b2289b6fec214fa88ee75a18d45c18f3.
Because in this study I analyze data from an already existing large-scale data set, the preregistration was done prior to accessing the data.
The preregistration was designed on the basis of the panel documentation online [@kittelAustrianCoronaPanel2020].
In some cases, it was impossible to execute the analyses as I had originally planned, for example because some properties of the variables only became apparent when seeing the actual data.
The most relevant deviations are reported below, and a complete list of all changes can be found in the online [companion website](https://XMtRA.github.io/Austrian_Corona_Panel_Project) (https://XMtRA.github.io/Austrian_Corona_Panel_Project).
